---
title: "R Problems"
output: html_document
---

Take the numbers 1 to 100, square them, and add all the even numbers while subtracting the odd ones!

```{r}
numbers <- c(1:100)
```

Define functions that will do what is needed, then sapply(numbers, function)

```{r}
is.even <- function(numbers) {
  numbers %% 2 == 0
}
```

```{r}
sq_numbers <- numbers^2
```

When you use brackets [], it only lets numbers through that are TRUE

```{r}
even_numbers <- sq_numbers[is.even(numbers)]
even_numbers
```

```{r}
odd_numbers <- sq_numbers[!is.even(numbers)]
odd_numbers
```

```{r}
answer <- sum(even_numbers) - sum(odd_numbers)
answer
```

You can also use `which()` to see which are TRUE

```{r}
which(is.even(numbers))
which(!is.even(numbers))
```

Scalers = numbers
Vectors = series of numbers

To make a Matrix, you can use the `cbind()` function

```{r}
n <- 25
tt <- seq(0,3.4,len=n)
x <- cbind(x1=tt, x2=tt^2)
head(x)
dim(x)
```

To create a matrix with 100 rows and 10 columns

```{r}
x <- matrix(1:1000, 100, 10)
head(x)
```
To return a specific entry, use []. [25,3] will return the 25th row and the 3rd column
```{r}
x[25,3]
```

Using the function cbind(), create a 10 x 5 matrix with first column x=1:10. Then use `2*x, 3*x, 4*x and 5*x` respectively in columns 2 through 5.

What is the sum of the elements of the 7th row?

```{r}
x = 1:10
X=cbind(x, 2*x, 3*x, 4*x, 5*x)
sum(X[7,])
```

Matrix Operations

